---
title: "Shared axis labels in patchwork plots"
description: |
  Getting shared axis labels in {patchwork} plots.

date: 2022-12-22

categories:
  - .Visualize
  - "{ggplot2}"
  - "{patchwork}"

params:
  use_renv: FALSE
  data_file: NULL # Filename of data located in /data/
  fair_dealing: FALSE
---

```{r setup, include=FALSE}
# Create templates for knitr code chunks --------------------------------------
knitr::opts_template$set(
  no.message = list(message = FALSE, warning = FALSE)
)
```

## Prerequisites

To access the datasets, help pages, and functions that we will use in this code snippet, load the following packages:

```{r prerequisite-packages, opts.label="no.message"}
library(ggplot2)
library(patchwork)
```

Then make some data and **ggplot2** plots to be used in the patchwork.

```{r}
huron <- data.frame(year = 1875:1972, level = as.vector(LakeHuron))
h <- ggplot(huron, aes(year))

h1 <- h +
  geom_ribbon(aes(ymin = level - 1, ymax = level + 1), fill = "grey70") +
  geom_line(aes(y = level))

h2 <- h + geom_area(aes(y = level))
```

## Shared x-axis labels

We set the bottom margin to 0 so the tag is in the same vertical position that the x-axis would otherwise be in.

```{r}
# Create the patchwork, dropping the x-axis labels from the plots, and setting
# the margins
h_patch <- h1 + h2 & xlab(NULL) & theme(plot.margin = margin(5.5, 5.5, 0, 5.5))

# Use the tag label as an x-axis label
wrap_elements(panel = h_patch) +
  labs(tag = "year") +
  theme(
    plot.tag = element_text(size = rel(1)),
    plot.tag.position = "bottom"
  )
```

## Shared y-axis labels

We set the left margin to 0 so the tag is in the same horizontal position that the y-axis would otherwise be in.

```{r}
# Create the patchwork, dropping the y-axis labels from the plots, and setting
# the margins
h_patch <- h1 / h2 & ylab(NULL) & theme(plot.margin = margin(5.5, 5.5, 5.5, 0))

# Use the tag label as a y-axis label
wrap_elements(h_patch) +
  labs(tag = "level") +
  theme(
    plot.tag = element_text(size = rel(1), angle = 90),
    plot.tag.position = "left"
  )
```

## Shared axis labels without using patchwork

[Elio Campitelli](https://eliocamp.github.io) shared a solution [on Mastodon](https://mastodon.social/@eliocamp/109559635126510130) that accomplishes the same results as above, but without patchwork. It uses the magic tilde notation to create functions in the `data` argument of each geom that adds a grouping variable `var` that can be faceted on.

```{r}
h <- ggplot(huron, aes(year)) +
  geom_ribbon(
    data = ~ transform(.x, var = "a"),
    aes(ymin = level - 1, ymax = level + 1),
    fill = "grey70"
  ) +
  geom_line(data = ~ transform(.x, var = "a"), aes(y = level)) +
  geom_area(data = ~ transform(.x, var = "b"), aes(y = level)) +
  # Since we don't care about the facet strips here, we can remove them.
  theme(
    strip.text = element_blank(),
    strip.background = element_blank()
  )
```

Facet by rows for a shared x-axis.

```{r}
h +
  facet_wrap(vars(var), scales = "free_y")
```

Facet by columns for a shared y-axis.

```{r}
h +
  facet_wrap(vars(var), scales = "free_y", ncol = 1)
```

{{< include /_common/_appendix.qmd >}}
