---
draft: true
title: Anime
description: |
  A template description.
author:
  - name: Michael McCarthy 
    url: ../../about.html
    affiliation: 
date: '2021-06-15'
categories:
  - 
repository_url: https://github.com/mccarthy-m-g/tidytales
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
# Ensure the project root is set up correctly using {here} --------------------
here::i_am("_posts/yyyy-mm-dd_post-name/post_name.Rmd")

# Load general set up packages used in all posts ------------------------------
library(here)

# Load R scripts located in R/ ------------------------------------------------

# Set default options for knitr code chunks -----------------------------------
knitr::opts_chunk$set(comment = "#>",
                      echo = TRUE)

# Create templates for knitr code chunks --------------------------------------
knitr::opts_template$set(
  fig.graphic = list(echo = FALSE, out.width = "100%"),
  no.message = list(message = FALSE, warning = FALSE)
)
```

```{r meta, include=FALSE, results='asis'}
library(metathis)

meta() %>%
  meta_social(
    title = "Tidy Tales: ...",
    url = "https://tidytales.ca/posts/yyyy-mm-dd_post-name/",
    image = "https://tidytales.ca/posts/yyyy-mm-dd_post-name/images/post-name_twittercard.png",
    image_alt = "Tidy Tales logo",
    og_type = "website",
    og_author = "Michael McCarthy",
    twitter_card_type = "summary_large_image",
    twitter_creator = "@mccarthymg"
  )
```

## Overview

## R

```{r requirements-setup}

```

### Wrangle

### Visualize

## Final Graphic

```{r final-plot, echo=FALSE, preview=TRUE}

```

## Comments {.appendix}

<details open class="comment-section">
   <summary>
   </summary>
</details>

## Data Source {.appendix}

Download the data used in this article with `tidytuesdayR::tt_load("yyyy-mm-dd")`.

## Session Info {.appendix}

<details class="session-info">
   <summary>
   </summary>
```{r session-info, echo=FALSE}
sessionInfo()
```
</details>

## Fair Dealing {.appendix}

Any of the trademarks, service marks, collective marks, design rights or similar rights that are mentioned, used, or cited in this article are the property of their respective owners. They are used here as fair dealing for the purpose of education in accordance with [section 29 of the Copyright Act](https://laws-lois.justice.gc.ca/eng/acts/c-42/page-8.html#h-103270) and do not infringe copyright.
